Deep Gaussian Processes and Hierarchical Modeling with Functional Data: A Case Study in Cosmological Power Spectra

### 2024-05-16 paper/main.tex
add Data section draft

The Cosmic Emu emulator is constructed by (process convolutions, ...). Segue to our method.

Due to the nonstationary behavior expected to be seen in power spectra (e.g., from baryonic acoustic oscillations), we expand on the previous emulation methods and propose a novel synthesis of methods to quantify uncertainty and predict matter power spectra for different cosmologies using deep Gaussian processes (DGPs). In order to handle multiple realizations of functional output (from the Mira-Titan simulation suite), we use hierarchical and basis functions modeling within our DGP framework to accurately estimate and predict power spectra for different cosmologies. When comparing to Cosmic Emu, we see favorable results for our method. The code to reproduce these results are available at \url{https://github.com/stevewalsh124/dgp.hm}.

\section{Data}

Here, we provide further details on the form of the data (in the introduction, we only briefly introduce it).  Include a figure of the raw data for a particular cosmology.  We describe the different modeling components and the regions of the space where they are trusted.  We also describe how each cosmology is a simulation based on 8 inputs, and the output $\mathcal{P}(k)$ is functional.

We will handle the data in two stages.  First, we will focus on an individual cosmology and obtain the best predicted curve.  Second, we will use these curves to predict the curve for a held-out cosmology.  These two stages form the main two sections of this paper.

Images:
\begin{itemize}
    \item Example cosmology showing 16 low res, 1 high res, 1 pert, wt avg, and where each output is deemed unbiased
\end{itemize}

\section{Bayesian Hierarchical Modeling for Particular Cosmologies}

@@

\subsection{Simulation Study}

@@

\subsection{Actual results (but more aptly named)}

@@

\section{Prediction for new cosmologies}

@@

\subsection{The model}

@@

\subsection{Actual results (but more aptly named)}

@@

\section{Discussion}

Brief recap, directions for future research: BSS-ANOVA in lieu of DGPs, hydrodynamical sims or TCs as other applications of this methodology. A more detailed sensitivity analysis for the parameters, modeling $W$ across different cosmologies.

\section{Appendix}

>>>>

\textcolor{blue}{The Cosmic Emu emulator is constructed by (process convolutions, ...). Segue to our method after and compare/contrast with Cosmic Emu.}

Due to the nonstationary behavior expected to be seen in power spectra (e.g., from baryonic acoustic oscillations), we expand on the previous emulation methods and propose a novel synthesis of methods to quantify uncertainty and predict matter power spectra for different cosmologies using deep Gaussian processes (DGPs). In order to handle multiple realizations of functional output (from the Mira-Titan simulation suite), we use hierarchical and basis functions modeling within our DGP framework to accurately estimate and predict power spectra for different cosmologies. When comparing to Cosmic Emu, we see favorable results for our method. The code to reproduce these results are available at \url{https://github.com/stevewalsh124/dgp.hm}. \textcolor{blue}{Please note: this repository is currently private until we hear back regarding permissions for sharing the Mira-Titan data used to train and predict this model.}

The remainder of the paper is organized as follows: Section \ref{sec:data} provides more details regarding the Mira-Titan simulation suite. Section \ref{sec:hm_fit} describes our model and the training procedure for both toy data as well as the Mira-Titan dataset. With this trained model, Section \ref{sec:pred} explains our modeling approach for predictions at held-out cosmologies and compares with other methods. Section \ref{sec:disc} concludes with a discussion of our contributions and avenues for future research.

\section{Data}
\label{sec:data}

The Mira-Titan simulation suite is a dataset consisting of simulated power spectra for 117 different cosmologies. Each of these cosmologies is specified by 8 cosmological parameter values, which we will denote as $\theta \in \mathbb{R}^8$. Different simulation output exists for different values of a ninth value, the redshift; in this work, we fix this value at 0. For each cosmology, we have a set of 18 simulated spectra: an inexpensive spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from lower-resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from a high resolution simulation ($y_h$).

These spectra are represented as a function of wavenumber $k$, and hence is denoted $P(k)$. Following \cite{moran2023mira}, we model on the emulation space $\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$. The output is available for $0.001 \leq k \leq 5$ for $n=351$ different values of $k$, but each data type has differing values where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable estimate of the spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and $y_h$ is valid for $0.04 \leq k \leq 5$. Figure \ref{fig:plot_data} shows an example of the output for a particular cosmology.

\begin{figure}[ht]
    \centering
    \includegraphics[width=4in]{plot_data.png}
    \caption{The perturbation theory, low resolution runs, and high resolution run for the first training cosmology. The weighted average is also shown as a dashed line. Dotted lines at the bottom indicate where each data type is deemed approximately unbiased.}
    \label{fig:plot_data}
\end{figure}

From the multiple low- and high-resolution spectra across all cosmologies, \cite{moran2023mira} obtain estimates for the precision of these spectra across the $n$ wavenumber values ($p_1,\dots,p_n$) using a log-log regression model, and a multiplier $c$ for the increase in precision from the low- to high- resolution output. Given that the perturbation theory spectra is non-stochastic, we treat this as having a precision of $10^6$. We leverage these precision estimates within our model and calculate a weighted average for each cosmology: $\bar y = AddThisIn$, \textcolor{blue}{where we have four diagonal covariance matrices.}

We will handle the data in two stages.  First, we will focus on 111 training cosmologies and obtain the best esimated spectrum for each.  Second, we will use these estimates to predict the spectra for six held-out cosmologies.

\section{Bayesian Hierarchical Modeling for Particular Cosmologies}
\label{sec:hm_fit}

@@

\subsection{Simulation Study}
\label{subsec:sim}

@@

\subsection{Estimating Spectra with Mira-Titan Data}
\label{subsec:mira_fit}

@@

\section{Prediction for Unobserved Cosmologies}
\label{sec:pred}

@@

\subsection{PCA-GP Model}
\label{subsec:pca}

@@

\subsection{Predicting Spectra from Mira-Titan Data}
\label{subsec:mira_pred}

@@

\section{Discussion}
\label{sec:disc}

Brief recap, directions for future research: BSS-ANOVA in lieu of DGPs, hydrodynamical sims or TCs as other applications of this methodology. A more detailed sensitivity analysis for the parameters, modeling $W$ across different cosmologies, consider modeling for different redshift values other than $z=0$.

\section{Appendix}
\label{sec:apdx}
